Global code motion of congruent instructions

### 2017-04-17 paper.tex
fuse section 4 into 3.5 codegen

\subsection{Code generation}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed, and the \SSA{} form is restored by updating the
intermediate representation (IR) to reflect the changes. At the same time
\MemorySSA{} is also updated to get up-to-date information about memory references.

@@

\section{\SSA{} based global code motion of identical computations}
\label{sec:implementation-details}
% Move this before legality and profitability??
% Ranking of instructions from Zadeck
% Liveness analysis from Ramakrishna
In this section we will present an overview of the algorithm we implemented in
\LLVM{}.  The amount of hoisting depends on whether we collect \GVN{} of
instructions before finding candidates (optimistic) or, on-demand
(pessimistic). It also depends on the generality of the \GVN{} algorithm as
mentioned earlier in Section~\ref{subsec:finding-candidates}. We have
implemented a optimistic global code motion of congruent instructions which uses
the liveness analysis as illustrated in Das \cite{das2012} and ranking
expressions explained in Rosen \cite{rosen1988global}.

Code-motion basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

\subsection{Optimistic code motion algorithm}
\label{subsec:optimistic}
We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithmn would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators. However,
it should be noted that we only hoist very busy expressions to avoid checking
for undefined behaviors resulting because of introducing extra computation in an
execution path.

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot not guarantee progress is marked
as a barrier. In the absense of context (as in our current implementation), some
instructions which might be safe are still classified a barrier, e.g., volatile
loads/stores, calls with missing attributes.

>>>>

\subsection{Code generation for the optimistic code motion algorithm}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting or sinking. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed, and the \SSA{} form is restored by
updating the intermediate representation (IR) to reflect the changes. At the
same time \MemorySSA{} is also updated to get up-to-date information about
memory references.

@@

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm as mentioned earlier in
Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
global code motion of congruent instructions which uses the liveness analysis as
illustrated in Das \cite{das2012} and ranking expressions explained in Rosen
\cite{rosen1988global}.

Code-motion basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining.) The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithmn would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators. However,
it should be noted that we only hoist very busy expressions to avoid checking
for undefined behaviors resulting because of introducing extra computation in an
execution path.

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot guarantee progress is marked
as a barrier. In the absence of context (as in our current implementation), some
instructions which might be safe are still classified as barriers, e.g., volatile
loads/stores, calls with missing attributes.
